---
title: "Urolithin A results"
author: Mia Nahom
date: 2025-2-4
format: 
  html:
     df-print: paged
     embed-resources: true
---

## Load libraries
```{r}
#| output: false
library("DESeq2")
library("apeglm")
library("pheatmap")
library("tidyverse")
library("ggrepel")
library("ashr")
library("goseq")
library("biomaRt")
library("ggplot2")
library("clusterProfiler")
library("enrichplot")
library("ggupset")
library("data.table")
```

## Read in distal and proximal objects and set working directory

```{r}
#setwd(dir="UrolithinA/")
directory <- "../counts/"
list.files(directory)

#Load Distal Table
DistalTable <- readRDS("../QC_meta/DistalTable.rds")
DistalTable
ProximalTable <- readRDS("../QC_meta/ProximalTable.rds")
ProximalTable
```
## Create DESeq2 Object
```{r}
distal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = DistalTable, 
  directory = directory, 
  design = ~ uroA_bin
)
proximal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = ProximalTable, 
  directory = directory, 
  design = ~ uroA_bin
)
```
## Distal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(distal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
distal_ddsHTSeq <- distal_ddsHTSeq[keep,]
```

### Set factors
```{r}
levels <- c("zero","low","low.med","med.high","high")
distal_ddsHTSeq$uroA_bin <- factor(distal_ddsHTSeq$uroA_bin, levels = levels)
distal_ddsHTSeq$uroA_bin
```
### Get results
```{r}
distal_dds <- DESeq(distal_ddsHTSeq)
resultsNames(distal_dds)
```

### Extract Results
#### res1: high vs zero
#### res2: med.high vs zero
#### res3: low.med vs zero
#### res4: low vs zero
#### res5: high vs low
#### res6: med.high vs low
#### res7: low.med vs low
#### res8: high vs low.med
#### res9: high.med vs low.med
#### res10: high vs med.high

```{r}
res1 <-results(distal_dds, name="uroA_bin_high_vs_zero")
res2 <-results(distal_dds, name="uroA_bin_med.high_vs_zero")
res3 <-results(distal_dds, name="uroA_bin_low.med_vs_zero")
res4 <-results(distal_dds, name="uroA_bin_low_vs_zero")
res5 <-results(distal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_low_vs_zero")))
res6 <-results(distal_dds, contrast=list(c("uroA_bin_med.high_vs_zero"),c("uroA_bin_low_vs_zero")))
res7 <-results(distal_dds, contrast=list(c("uroA_bin_low.med_vs_zero"),c("uroA_bin_low_vs_zero")))
res8 <-results(distal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_low.med_vs_zero")))
res9 <-results(distal_dds, contrast=list(c("uroA_bin_med.high_vs_zero"),c("uroA_bin_low.med_vs_zero")))
res10 <-results(distal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_med.high_vs_zero")))
```
### Summary of results
```{r}
summary(res1)
summary(res2)
summary(res3)
summary(res4)
summary(res5)
summary(res6)
summary(res7)
summary(res8)
summary(res9)
summary(res10)
```

## Proximal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(proximal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
proximal_ddsHTSeq <- proximal_ddsHTSeq[keep,]
```
### Set factors
```{r}
levels <- c("zero","low","low.med","med.high","high")
proximal_ddsHTSeq$uroA_bin <- factor(proximal_ddsHTSeq$uroA_bin, levels = levels)
proximal_ddsHTSeq$uroA_bin
```
### Get results
```{r}
proximal_dds <- DESeq(proximal_ddsHTSeq)
resultsNames(proximal_dds)
```
### Extract Results
#### res1: high vs zero
#### res2: med.high vs zero
#### res3: low.med vs zero
#### res4: low vs zero
#### res5: high vs low
#### res6: med.high vs low
#### res7: low.med vs low
#### res8: high vs low.med
#### res9: high.med vs low.med
#### res10: high vs med.high

```{r}
res1 <-results(proximal_dds, name="uroA_bin_high_vs_zero")
res2 <-results(proximal_dds, name="uroA_bin_med.high_vs_zero")
res3 <-results(proximal_dds, name="uroA_bin_low.med_vs_zero")
res4 <-results(proximal_dds, name="uroA_bin_low_vs_zero")
res5 <-results(proximal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_low_vs_zero")))
res6 <-results(proximal_dds, contrast=list(c("uroA_bin_med.high_vs_zero"),c("uroA_bin_low_vs_zero")))
res7 <-results(proximal_dds, contrast=list(c("uroA_bin_low.med_vs_zero"),c("uroA_bin_low_vs_zero")))
res8 <-results(proximal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_low.med_vs_zero")))
res9 <-results(proximal_dds, contrast=list(c("uroA_bin_med.high_vs_zero"),c("uroA_bin_low.med_vs_zero")))
res10 <-results(proximal_dds, contrast=list(c("uroA_bin_high_vs_zero"),c("uroA_bin_med.high_vs_zero")))
```
### Summary of results
```{r}
summary(res1)
summary(res2)
summary(res3)
summary(res4)
summary(res5)
summary(res6)
summary(res7)
summary(res8)
summary(res9)
summary(res10)
```