---
title: "Urolithin M7 results"
author: Mia Nahom
date: 2025-2-4
format: 
  html:
     df-print: paged
     embed-resources: true
---

## Load libraries
```{r}
#| output: false
library("DESeq2")
library("apeglm")
library("pheatmap")
library("tidyverse")
library("ggrepel")
library("ashr")
library("goseq")
library("biomaRt")
library("ggplot2")
library("clusterProfiler")
library("enrichplot")
library("ggupset")
library("data.table")
library("dplyr")
```

## Read in distal and proximal objects and set working directory

```{r}
#setwd(dir="../UrolithinM7/")
directory <- "../counts/"
list.files(directory)

#Load Distal Table
DistalTable <- readRDS("../QC_meta/DistalTable.rds")
DistalTable
ProximalTable <- readRDS("../QC_meta/ProximalTable.rds")
ProximalTable
annotation <- readRDS("../QC_meta/annotation.rds")
annotation
```
## Create DESeq2 Object
```{r}
distal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = DistalTable, 
  directory = directory, 
  design = ~ uroM7_bin
)
proximal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = ProximalTable, 
  directory = directory, 
  design = ~ uroM7_bin
)
```
## Distal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(distal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
distal_ddsHTSeq <- distal_ddsHTSeq[keep,]
```

### Set factors
```{r}
levels <- c("zero","low","med","high")
distal_ddsHTSeq$uroM7_bin <- factor(distal_ddsHTSeq$uroM7_bin, levels = levels)
distal_ddsHTSeq$uroM7_bin
```
### Get results
```{r}
distal_dds <- DESeq(distal_ddsHTSeq)
resultsNames(distal_dds)
```

### Extract Results
#### res1: high vs zero
#### res2: med vs zero
#### res3: low vs zero
#### res4: high vs low
#### res5: med vs low
#### res6: high vs med

```{r}
res1_dist <-results(distal_dds, name="uroM7_bin_high_vs_zero")
res2_dist <-results(distal_dds, name="uroM7_bin_med_vs_zero")
res3_dist <-results(distal_dds, name="uroM7_bin_low_vs_zero")
res4_dist <-results(distal_dds, contrast=list(c("uroM7_bin_high_vs_zero"),c("uroM7_bin_low_vs_zero")))
res5_dist <-results(distal_dds, contrast=list(c("uroM7_bin_med_vs_zero"),c("uroM7_bin_low_vs_zero")))
res6_dist <-results(distal_dds, contrast=list(c("uroM7_bin_high_vs_zero"),c("uroM7_bin_med_vs_zero")))

```
### Summary of results
```{r}
summary(res1_dist)
summary(res2_dist)
summary(res3_dist)
summary(res4_dist)
summary(res5_dist)
summary(res6_dist)

```
```{r}

res1_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res1_dist.csv")

res2_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res2_dist.csv")

res3_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res3_dist.csv")

res4_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res4_dist.csv")

res5_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res5_dist.csv")

res6_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res6_dist.csv")
```

## Proximal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(proximal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
proximal_ddsHTSeq <- proximal_ddsHTSeq[keep,]
```
### Set factors
```{r}
levels <- c("zero","low","med","high")
proximal_ddsHTSeq$uroM7_bin <- factor(proximal_ddsHTSeq$uroM7_bin, levels = levels)
proximal_ddsHTSeq$uroM7_bin
```
### Get results
```{r}
proximal_dds <- DESeq(proximal_ddsHTSeq)
resultsNames(proximal_dds)
```

### Extract Results
#### res1: high vs zero
#### res2: med vs zero
#### res3: low vs zero
#### res4: high vs low
#### res5: med vs low
#### res6: high vs med

```{r}
res1_prox <-results(proximal_dds, name="uroM7_bin_high_vs_zero")
res2_prox <-results(proximal_dds, name="uroM7_bin_med_vs_zero")
res3_prox <-results(proximal_dds, name="uroM7_bin_low_vs_zero")
res4_prox <-results(proximal_dds, contrast=list(c("uroM7_bin_high_vs_zero"),c("uroM7_bin_low_vs_zero")))
res5_prox <-results(proximal_dds, contrast=list(c("uroM7_bin_med_vs_zero"),c("uroM7_bin_low_vs_zero")))
res6_prox <-results(proximal_dds, contrast=list(c("uroM7_bin_high_vs_zero"),c("uroM7_bin_med_vs_zero")))

```
### Summary of results
```{r}
summary(res1_prox)
summary(res2_prox)
summary(res3_prox)
summary(res4_prox)
summary(res5_prox)
summary(res6_prox)
```
```{r}
res1_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res1_prox.csv")

res2_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res2_prox.csv")

res3_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res3_prox.csv")

res4_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res4_prox.csv")

res5_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res5_prox.csv")

res6_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res6_prox.csv")
```


### Look at some counts
```{r}
res5_prox %>% as.data.frame() %>% arrange(padj) %>% head()
```
#### Plot Counts

```{r}
plotCounts(proximal_dds,gene="ENSG00000187908", intgroup = "uroM7_bin")
plotCounts(proximal_dds,gene="ENSG00000165238", intgroup = "uroM7_bin")
plotCounts(proximal_dds,gene="ENSG00000178199", intgroup = "uroM7_bin")
plotCounts(proximal_dds,gene="ENSG00000183665", intgroup = "uroM7_bin")
plotCounts(proximal_dds,gene="ENSG00000162878", intgroup = "uroM7_bin")
```