---
title: "Isorolithin A results"
author: Mia Nahom
date: 2025-2-4
format: 
  html:
     df-print: paged
     embed-resources: true
---

## Load libraries
```{r}
#| output: false
library("DESeq2")
library("apeglm")
library("pheatmap")
library("tidyverse")
library("ggrepel")
library("ashr")
library("goseq")
library("biomaRt")
library("ggplot2")
library("clusterProfiler")
library("enrichplot")
library("ggupset")
library("data.table")
```

## Read in distal and proximal objects and set working directory

```{r}
#setwd(dir="../IsourolithinA/")
directory <- "../counts/"
list.files(directory)

#Load Distal Table
DistalTable <- readRDS("../QC_meta/DistalTable.rds")
DistalTable
ProximalTable <- readRDS("../QC_meta/ProximalTable.rds")
ProximalTable
annotation <- readRDS("../QC_meta/annotation.rds")
annotation
```
## Create DESeq2 Object
```{r}
distal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = DistalTable, 
  directory = directory, 
  design = ~ isouroA_bin
)
proximal_ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = ProximalTable, 
  directory = directory, 
  design = ~ isouroA_bin
)
```
## Distal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(distal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
distal_ddsHTSeq <- distal_ddsHTSeq[keep,]
```

### Set factors
```{r}
levels <- c("zero","low","low.med","med.high","high")
distal_ddsHTSeq$isouroA_bin <- factor(distal_ddsHTSeq$isouroA_bin, levels = levels)
distal_ddsHTSeq$isouroA_bin
```
### Get results
```{r}
distal_dds <- DESeq(distal_ddsHTSeq)
resultsNames(distal_dds)
```

### Extract Results
#### res1: high vs zero
#### res2: med.high vs zero
#### res3: low.med vs zero
#### res4: low vs zero
#### res5: high vs low
#### res6: med.high vs low
#### res7: low.med vs low
#### res8: high vs low.med
#### res9: high.med vs low.med
#### res10: high vs med.high

```{r}
res1_dist <-results(distal_dds, name="isouroA_bin_high_vs_zero")
res2_dist <-results(distal_dds, name="isouroA_bin_med.high_vs_zero")
res3_dist <-results(distal_dds, name="isouroA_bin_low.med_vs_zero")
res4_dist <-results(distal_dds, name="isouroA_bin_low_vs_zero")
res5_dist <-results(distal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_low_vs_zero")))
res6_dist <-results(distal_dds, contrast=list(c("isouroA_bin_med.high_vs_zero"),c("isouroA_bin_low_vs_zero")))
res7_dist <-results(distal_dds, contrast=list(c("isouroA_bin_low.med_vs_zero"),c("isouroA_bin_low_vs_zero")))
res8_dist <-results(distal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_low.med_vs_zero")))
res9_dist <-results(distal_dds, contrast=list(c("isouroA_bin_med.high_vs_zero"),c("isouroA_bin_low.med_vs_zero")))
res10_dist <-results(distal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_med.high_vs_zero")))
```
### Summary of results
```{r}
summary(res1_dist)
summary(res2_dist)
summary(res3_dist)
summary(res4_dist)
summary(res5_dist)
summary(res6_dist)
summary(res7_dist)
summary(res8_dist)
summary(res9_dist)
summary(res10_dist)
```
```{r}
res1_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res1_dist.csv")

res2_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res2_dist.csv")

res3_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res3_dist.csv")

res4_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res4_dist.csv")

res5_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res5_dist.csv")

res6_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res6_dist.csv")


res7_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res7_dist.csv")

res8_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res8_dist.csv")

res9_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res9_dist.csv")

res10_dist %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res10_dist.csv")
```

## Proximal Analysis
### Filter out low expression genes
```{r}
# sum counts for each gene across samples
sumcounts <- rowSums(counts(proximal_ddsHTSeq))
# take the log
logsumcounts <- log(sumcounts,base=10)
# plot a histogram of the log scaled counts
hist(logsumcounts,breaks=100)
# get genes with summed counts greater than 50
keep <- sumcounts > 50
proximal_ddsHTSeq <- proximal_ddsHTSeq[keep,]
```
### Set factors
```{r}
levels <- c("zero","low","low.med","med.high","high")
proximal_ddsHTSeq$isouroA_bin <- factor(proximal_ddsHTSeq$isouroA_bin, levels = levels)
proximal_ddsHTSeq$isouroA_bin
```
### Get results
```{r}
proximal_dds <- DESeq(proximal_ddsHTSeq)
resultsNames(proximal_dds)
```
### Extract Results
#### res1: high vs zero
#### res2: med.high vs zero
#### res3: low.med vs zero
#### res4: low vs zero
#### res5: high vs low
#### res6: med.high vs low
#### res7: low.med vs low
#### res8: high vs low.med
#### res9: high.med vs low.med
#### res10: high vs med.high

```{r}
res1_prox <-results(proximal_dds, name="isouroA_bin_high_vs_zero")
res2_prox <-results(proximal_dds, name="isouroA_bin_med.high_vs_zero")
res3_prox <-results(proximal_dds, name="isouroA_bin_low.med_vs_zero")
res4_prox <-results(proximal_dds, name="isouroA_bin_low_vs_zero")
res5_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_low_vs_zero")))
res6_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_med.high_vs_zero"),c("isouroA_bin_low_vs_zero")))
res7_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_low.med_vs_zero"),c("isouroA_bin_low_vs_zero")))
res8_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_low.med_vs_zero")))
res9_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_med.high_vs_zero"),c("isouroA_bin_low.med_vs_zero")))
res10_prox <-results(proximal_dds, contrast=list(c("isouroA_bin_high_vs_zero"),c("isouroA_bin_med.high_vs_zero")))
```
### Summary of results
```{r}
summary(res1_prox)
summary(res2_prox)
summary(res3_prox)
summary(res4_prox)
summary(res5_prox)
summary(res6_prox)
summary(res7_prox)
summary(res8_prox)
summary(res9_prox)
summary(res10_prox)
```

```{r}
res1_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res1_prox.csv")

res2_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res2_prox.csv")

res3_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res3_prox.csv")

res4_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res4_prox.csv")

res5_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res5_prox.csv")

res6_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res6_prox.csv")

res7_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res7_prox.csv")

res8_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res8_prox.csv")

res9_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res9_prox.csv")

res10_prox %>%
  as.data.frame() %>%
  arrange(padj) %>%
  tibble::rownames_to_column(var = "ensembl_gene_id") %>%
  merge(annotation, by.x = "ensembl_gene_id", by.y = "ensembl_gene_id", all.x = TRUE) %>%
  write.csv(.,file="res10_prox.csv")
```

